Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-08-23 analysis/paper/paper.qmd
some more eidts on methods

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial centers were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Yilan, located in northeastern Taiwan, is an example of a periphery region where European control was limited and indirect for most of the time but intensively involved in regional trade activities stimulated by the European presence. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export on a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017]. Despite some discussions on the operations and policies of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people with the commercial activities are rare. The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jorden2009]. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

@@

# Mateirals and Methods
To explore changes in resource use before and after the European arrival with associated long-distance trade, I analyzed zooarchaeological and botanical assemblages from 40 adjacent units at the residential area excavated during the first season (2001-2004) at Kiwulan [@Chen2007]. These 40 units were selected based on their relatively intact archaeological contexts and spatial continuity, providing a consistent framework for understanding daily food practices throughout different time periods on a settlement scale.

To construct a reliable chronology, the archaeological records were divided into six temporal layers spanning from the 14th century to the 19th century based on excavation depth, stratigraphic data, sediment color, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above still provides a feasible framework for analyzing archaeological records before and after the European influence. Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with Chinese migrants.

In this study, only the faunal assemblages identified to at least the family level were considered for consistent comparison across time sequences. To obtain a general pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. In this analysis, ubiquity values are expressed as the percentage of units with a given taxon to the total number of units containing cultural layers. Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method for a broad overview of taxa through time [@Grayson2004; @Lyman2012]. For botanical assemblages, their number was served as supplementary evidence for food practices due to potential biases that caused by differential preservation conditions. For example, carbonized seeds usually preserve better than uncarbonized ones, which could lead to a misleading interpretation.

>>>>

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial centers were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Yilan, located in northeastern Taiwan, is an example of a periphery region where European control was limited and indirect for most of the time but intensively involved in regional trade activities stimulated by the European presence. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export on a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017]. Despite some discussions on the operations and policies of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people with the commercial activities are rare. The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jordan2009]. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

@@

# Mateirals and Methods
To explore changes in resource use before and after the European arrival with associated long-distance trade, I examined zooarchaeological and botanical remains from 40 adjacent units at the major residential area excavated during the first season (2001-2004) at Kiwulan [@Chen2007]. The remains considered in this study are from dwelling spaces and trash pits that reflect daily dietary practices, whereas those from burials that might relate to ritual activities are excluded. These 40 units were selected because of their relatively intact archaeological contexts and spatial continuity, providing a consistent framework for understanding daily food practices over time on a settlement scale. Moreover, the same excavation and recovery methods applied at the site allow comparison between assembalges from different temporal sequences.

To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six temporal layers spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above provides a robust framework for comparing archaeological records before and after the European influence. Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with the arrival of Chinese migrants in Yilan.

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units containing cultural layers. Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method [@Grayson2004; @Lyman2012]. Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size, and therefore is used throughout this study [@Grayson1984; @Lyman2008]. For botanical assemblages, their number was served as supplementary evidence for food practices due to potential biases that caused by differential preservation conditions. For example, carbonized seeds usually preserve better than uncarbonized ones, which could lead to a misleading interpretation.
